---
title: "S2S Lab 5"
subtitle: "Exercise Solutions"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, comment=NA)
library(PASWR2)
library(knitr)
library(kableExtra)
```

# Exercise 1 {-}

A university offers 8 different statistics courses to its honours level students. The numbers of students in each course are as follows:

* There are three course which have 85 students enrolled
* There are two course which have 112 students enrolled
* There is one course with 108 students enrolled
* There is one course with 95 students enrolled
* There is one course with only 46 students enrolled

Suppose you were to randomly sample three of these statistics courses **with** replacement. What would the sampling distribution of the sample mean number of students enrolled, $\bar X$, be? What would the sampling distribution of the sample variance in the number of students enrolled, $S^2$, be?

Can you use your sampling distributions to find $E[\bar X]$, $\sigma_{\bar X}^2$ and $E[S^2]$?

## Solution {-}

Create a vector `students` which contains the number of students enrolled in each of the 8 courses. This vector can be given as the arguments in `expand.grid()` to show all possible combinations of sampling three courses with replacement. There are 512 different combinations.

```{r}
students <- c(rep(x = 85, times = 3), rep(x = 112, times = 2), 108, 95, 46)

samples_w <- expand.grid(Class1 = students, Class2 = students, Class3 = students)
```

Calculate the sample mean and sample variance for each possible sample of 3 courses. Update `samples_w` to contain these values and the sample they came from.

```{r}
sample_mean_w <- apply(X = samples_w, MARGIN = 1, FUN = mean)

sample_var_w <- apply(X = samples_w, MARGIN = 1, FUN = var)

samples_w <- cbind(samples_w, Mean = sample_mean_w, Variance = sample_var_w)
```

Use `samples_w` to find the sampling distribution of $\bar X$.

```{r}
samplingdist_mean_w <- table(samples_w$Mean)/512
```

\newpage
```{r, echo=FALSE, fig.pos="H"}
sampdist_mean_table <- data.frame(means1 = round(sort(unique(samples_w$Mean))[1:12], 3),
                                  probs1 = round(as.numeric(samplingdist_mean_w[1:12]), 4),
                                  means2 = round(sort(unique(samples_w$Mean))[13:24], 3),
                                  probs2 = round(as.numeric(samplingdist_mean_w[13:24]), 4),
                                  means3 = c(round(sort(unique(samples_w$Mean))[25:35], 3), ""),
                                  probs3 = c(round(as.numeric(samplingdist_mean_w[25:35]), 4), ""))
names(sampdist_mean_table) <- rep(c("$\\bar x_i$", "$P(\\bar X=\\bar x_i)$"), times = 3)

kable(sampdist_mean_table,
      caption = "Sampling distribution for $\\bar X$ when sampling with replacement.",
      escape = FALSE) %>%
  column_spec(2, border_right = TRUE) %>%
  column_spec(4, border_right = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

```{r, echo=FALSE, include=FALSE}
means_w <- sort(unique(samples_w$Mean))
mean_Xbar_w <- sum(means_w*samplingdist_mean_w)
mean_Xbar_w
```

The expected value of $\bar X$ can be calculated as $E[\bar X]=`r mean_Xbar_w`$.

```{r}
means_w <- sort(unique(samples_w$Mean))
mean_Xbar_w <- sum(means_w*samplingdist_mean_w)
mean_Xbar_w
```

```{r, echo=FALSE, include=FALSE}
var_Xbar_w <- sum(means_w^2*samplingdist_mean_w)-(mean_Xbar_w)^2
var_Xbar_w
```

The variance of $\bar X$ can be calculated as $\sigma^2_{\bar X}=`r round(var_Xbar_w, 4)`$.

```{r}
var_Xbar_w <- sum(means_w^2*samplingdist_mean_w)-(mean_Xbar_w)^2
var_Xbar_w
```

Use `samples_w` to find the sampling distribution of $S^2$.

```{r}
samplingdist_var_w <- table(samples_w$Variance)/512
```

```{r, echo=FALSE, fig.pos="h"}
sampdist_var_table <-  data.frame(vars1 = sort(unique(round(samples_w$Variance, 3)))[1:7],
                                  probs1 = round(as.numeric(samplingdist_var_w[1:7]), 4),
                                  vars2 = sort(unique(round(samples_w$Variance, 3)))[8:14],
                                  probs2 = round(as.numeric(samplingdist_var_w[8:14]), 4),
                                  vars3 = sort(unique(round(samples_w$Variance, 3)))[15:21],
                                  probs3 = round(as.numeric(samplingdist_var_w[15:21]), 4))
names(sampdist_var_table) <- rep(c("$s^2_i$", "$P(S^2=s^2_i)$"), times = 3)

kable(sampdist_var_table,
      caption = "Sampling distribution for $S^2$ when sampling with replacement.",
      escape = FALSE) %>%
  column_spec(2, border_right = TRUE) %>%
  column_spec(4, border_right = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

```{r, echo=FALSE, include=FALSE}
vars_w <- sort(unique(round(samples_w$Variance, 4)))
mean_S2_w <- sum(vars_w*samplingdist_var_w)
mean_S2_w
```

The expected value of the sample variance can then be calculated as $E[S^2]=`r round(mean_S2_w)`$.

```{r}
vars_w <- sort(unique(round(samples_w$Variance, 4)))
mean_S2_w <- sum(vars_w*samplingdist_var_w)
mean_S2_w
```


\newpage
# Exercise 2 {-}

Now suppose you want to randomly sample three of the above statistics courses **without** replacement. Find the sampling distribution of the sample mean number of students, $\bar X$ and the sampling distribution of the sample variance in the number of students enrolled, $S^2$.

Use these sampling distributions to find $E[\bar X]$, $\sigma_{\bar X}^2$ and $E[S^2]$.

How do these values compare to those found in Exercise 1?

## Solution {-}

Load the package `PASWR2` because sampling without replacement is used, so to list all the combinations of three class sizes that can be drawn, the `srs()` function is needed.

```{r}
library(PASWR2)

samples_wo <- srs(popvalues = students, n = 3)
dimnames(samples_wo) <- list(NULL, c("Class1", "Class2", "Class3"))
```

Calculate the sample mean and sample variance for each possible sample of 3 courses. Update `samples_wo` to contain these values and the sample they came from.

```{r}
sample_wo_means <- apply(X = samples_wo, MARGIN = 1, FUN = mean)

sample_wo_variances <- apply(X = samples_wo, MARGIN = 1, FUN = var)

samples_wo <- data.frame(samples_wo, Mean = sample_wo_means, Variance = sample_wo_variances)
```

Use `samples_wo` to find the sampling distribution of $\bar X$.

```{r}
samplingdist_mean_wo <- table(samples_wo$Mean)/56
```

```{r, echo=FALSE, fig.pos="H"}
sampdistwo_mean_table <- data.frame(means1 = round(sort(unique(samples_wo$Mean))[1:10], 3),
                                  probs1 = round(as.numeric(samplingdist_mean_wo[1:10]), 4),
                                  means2 = c(round(sort(unique(samples_wo$Mean))[11:19], 3), ""),
                                  probs2 = c(round(as.numeric(samplingdist_mean_wo[11:19]), 4), ""))
names(sampdistwo_mean_table) <- rep(c("$\\bar x_i$", "$P(\\bar X=\\bar x_i)$"), times = 2)

kable(sampdistwo_mean_table,
      caption = "Sampling distribution for $\\bar X$ when sampling without replacement.",
      escape = FALSE) %>%
  column_spec(2, border_right = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

```{r, include=FALSE}
means_wo <- sort(unique(samples_wo$Mean))
mean_Xbar_wo <- sum(means_wo*samplingdist_mean_wo)
mean_Xbar_wo
```

The expected value of $\bar X$ can be calculated as $E[\bar X]=`r mean_Xbar_wo`$.

```{r}
means_wo <- sort(unique(samples_wo$Mean))
mean_Xbar_wo <- sum(means_wo*samplingdist_mean_wo)
mean_Xbar_wo
```

```{r, include=FALSE}
var_Xbar_wo <- sum(means_wo^2*samplingdist_mean_wo)-(mean_Xbar_wo)^2
var_Xbar_wo
```

The variance of $\bar X$ can be calculated as $\sigma^2_{\bar X}=`r round(var_Xbar_wo, 4)`$.

```{r}
var_Xbar_wo <- sum(means_wo^2*samplingdist_mean_wo)-(mean_Xbar_wo)^2
var_Xbar_wo
```

Use `samples_wo` to find the sampling distribution of $S^2$.

```{r}
samplingdist_var_wo <- table(samples_wo$Variance)/56
```

```{r, echo=FALSE, fig.pos="h"}
sampdistwo_var_table <-  data.frame(vars1 = sort(unique(round(samples_wo$Variance, 3)))[1:9],
                                  probs1 = round(as.numeric(samplingdist_var_wo[1:9]), 4),
                                  vars2 = sort(unique(round(samples_wo$Variance, 3)))[10:18],
                                  probs2 = round(as.numeric(samplingdist_var_wo[10:18]), 4))
names(sampdistwo_var_table) <- rep(c("$s^2_i$", "$P(S^2=s^2_i)$"), times = 2)

kable(sampdistwo_var_table,
      caption = "Sampling distribution for $S^2$ when sampling without replacement.",
      escape = FALSE) %>%
  column_spec(2, border_right = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

```{r, include=FALSE}
vars_wo <- sort(unique(round(samples_wo$Variance, 4)))
mean_S2_wo <- sum(vars_wo*samplingdist_var_wo)
mean_S2_wo
```

The expected value of the sample variance can then be calculated as $E[S^2]=`r round(mean_S2_wo, 4)`$.

```{r}
vars_wo <- sort(unique(round(samples_wo$Variance, 4)))
mean_S2_wo <- sum(vars_wo*samplingdist_var_wo)
mean_S2_wo
```









