# Functions {#functions}

We have already seen and used many R functions. Now we are going to learn about some more statistics specific functions, as well as creating our own functions which will give us a lot of flexibility with what we can do in R.

## Probability functions

There are a range of functions we can use which make working with statistical distributions a lot easier. We are able to generate random numbers from a distribution, calculate cumulative probabilities, compute densities and return quantiles with them.

The name of the distribution you are using is going to be part of the function name, so to keep things simple, let's start by looking at the normal distribution. The four functions you can use with the normal distribution are:

* `rnorm(n, mean = 0, sd = 1)`: this returns a random sample of size `n` from a $N\sim(0,\,1)$ distribution.
* `pnorm(q, mean = 0, sd = 1)`: this computes the probability $\mathbb{P}(X\leq q)$, where $X\sim N(0,\,1)$.
* `dnorm(x, mean = 0, sd = 1)`: this computes the value of the probability density function, $f(x)$.
* `qnorm(p, mean = 0, sd = 1)`: this computes the quantile $x$ such that $\mathbb{P}(X\leq x)=p$, where $X\sim N(0,\,1)$.

Note that `norm` in each function name is because we are using the normal distribution. This part of the function names can be changed based on the distribution we want to use.

The arguments `mean = 0` and `sd = 1` are the default values, so leaving them out of any of the functions above means you will be using the $N(0,\,1)$ distribution. You can change the mean or standard deviation of the distribution by changing the values assigned to the arguments.

The values that `pnorm(q)`, `qnorm(p)` and `dnorm(x)` return can be summarised graphically in Figure \@ref(fig:normal-image).

```{r normal-image, echo=FALSE, fig.cap="Summary of probability functions for the standard normal distribution.", out.width="1200px", fig.pos="H", fig.align="center"}
knitr::include_graphics("Images/normal_plots.svg", dpi=400)
```

For example, we know that the 97.5^th^ quantile of the $N(0,\,1)$ distribution is roughly 1.96. We can double check this using the following code.

```{r func1}
qnorm(0.975)
```

::: {.webex-box .webex-check}

```{r func-ans1, echo=FALSE}
opts1 <- c("`rnorm`", "`pnorm`", "`dnorm`", answer="`qnorm`")
opts2 <- c("`rnorm`", "`pnorm`", "`dnorm`", answer="`qnorm`")
opts3 <- c("`rnorm`", answer="`pnorm`", "`dnorm`", "`qnorm`")
```


**Choose the correct function and complete the code for the following scenarios.**

* You want to construct a 90% confidence interval so need to know the 95^th^ quantile of the standard normal distribution.

     `r mcq(opts1)``(``r fitb(c("0.9", "0.90", "0.900", "p=0.9", "p=0.90", "p=0.900"))``, mean = 0, sd = 1)`

* How would you find the value of $x$ such that $\mathbb{P}(X\leq x)=0.45$, where $X\sim N(100,\,4^2)$?

     `r mcq(opts2)``(0.45, ``r fitb(c("mean = 100, sd=4", "100, 4", "sd=4, mean=100", "mean=100, 4", "100, sd=4"))``)`
     
* You want to know the proportion of the $N(0,\,2^2)$ distribution that lies below -2. That is, $\mathbb{P}(X\leq -2)$, where $X\sim N(0,\,2^2)$.

     `r mcq(opts3)``(``r fitb(c(-2, "q=-2"))``, sd =``r fitb(2)``)`

:::

The normal distribution is not the only one that we can use the four functions introduced above with. Table \@ref(tab:dist-tab) shows some other distributions that can be used. In order to use a different distribution, simply change the `norm` part in the function name to the distribution's R name. Change the arguments `mean =` and `sd =` to the relevant arguments for your chosen distribution as well.

```{r dist-tab-setup, echo=FALSE}
dist <- c("Normal", "Binomial", "Exponential", "Geometric", "Hypergeometric", "Negative Binomial",
          "Poisson", "Student's t", "Uniform", "Chi-square")
func_names <- c("`norm`", "`binom`", "`exp`", "`geom`", "`hyper`", "`nbinom`", "`pois`", "`t`", 
                "`unif`", "`chisq`")
args <- c("`mean = 0`: the mean with default value 0.\\\n`sd = 1`: the standard deviation with default value 1.",
          "`size =`: the number of trials.\\\n`prob =`: the probability of success for each trial.",
          "`rate = 1`: the value of $\\theta$ with default value 1.",
          "`prob =`: the probability of success in each trial.",
          "`m =`: the number of objects of type I in the population.\\\n`n =`: the number of objects **not** of type I in the population.\\\n`k =`: the size of the sample taken from the population.",
          "`size =`: the number of successful trials you want to observe.\\\n`prob =`: the probability of success in each trial.",
          "`lambda =`: the value of $\\lambda$.",
          "`df =`: the degrees of freedom.",
          "`min = 0`: the lower limit of the distribution with default value 0.\\\n`max = 1`: the upper limit of the distribution with default value 1.",
          "`df =`: the degrees of freedom.")

dist_table <- data.frame(Distribution = dist, `R Name` = func_names, Arguments = args)
colnames(dist_table)[2] <- c("R Name")
```

```{r dist-tab, echo=FALSE}
kable(dist_table, booktabs=TRUE, caption="Distribution names and arguments to use with probability functions.") %>%
  column_spec(1:3, background="white")
```

<!-- exercise using other probability distributions -->

---

To read more, see [Section 1.15 Probability Functions](https://read.kortext.com/reader/pdf/92741/68) in *Probability and Statistics with R*.

## Flow control

Flow control is the term used to describe the order that your code is carried out. Usually this just happens line by line, but there may be cases where you want to repeat and update certain code over and over. To save you time having to type out a lot of repeating code, there are functions in R which can help you to repeat certain operations.

The `for()` function allows you to repeat lines of code for a given number of repetitions. Sometimes using the `for()` function is referred to as a **for loop** because once you have run multiple lines of code, it loops back to the beginning and does it all again. The setup is:

```{r func2, eval=FALSE}
for(name in vector){
  statements
}
```

* `name`: this is the name you want to use for the index in each iteration of the repeated statements. Most commonly this is given the value `i`.
* `vector`: this is a vector which is the same length as the number of times you want to repeat the statements.
* `statements`: these are the lines of code you want to repeat a number of times.

In order to fully understand how the `for()` function works, let's look at an example. We could use the `for()` function to print the numbers 1 up to 5.

```{r func3}
for(i in 1:5){
  print(i)
}
```
Here, we have assigned the value `i` to the `name` argument and the vector is `1:5`. The `:` operator returns the sequence of numbers from 1 to 5 in this case.

The for loop will first of all give `i` the value 1 and then run the statements within `{ }`, so here it will run `print(1)`. `print()` is a function we haven't seen yet, but it just 'prints' the arguments to the R console so we will see the value 1 written there.

Next, the for loop will update the value of `i` to the next one given in the `vector` argument i.e. it will assign the value 2 to `i`. Then it will again run `print(2)` and update the value of `i` again. This will repeat until `i` has taken the all the values given in the `vector` argument.

::: {.webex-box}

**The following code has been partially written to sum together the numbers 1 to 12. Copy the code into your own R script and fill in each `__` to complete it.**

```{r func-ans2.1, eval=FALSE}
sum <- __

for(i in __){
  sum <- sum + __
}
```

`r hide("Solution")`
We start by creating the vector `sum` to which each value 1, 2, ..., 12 can be added. Initially it needs to take the value 0.

Within the function `for()`, we want `i` to, in turn, take each value 1, 2, ..., 12, so we need to provide a vector of these values (`1:12`). `sum` should then be updated each time `i` takes a new value, by adding it on to the old value of `sum`.

```{r func-ans2.2}
sum <- 0

for(i in 1:12){
  sum <- sum + i
}
```

This for loop starts with `sum` having the value 0. It will first assign 1 to `i` and execute the code `sum <- 0 + 1`, meaning `sum` now has the value 1.

* `i` will then be updated to take the value 2 and the for loop will run the code `sum <- 1 + 2` i.e. `sum` has the value 3.

* `i` will then be updated to take the value 3 and the for loop will run the code `sum <- 3 + 3` i.e. `sum` has the value 6.

* $\vdots$

This repeats until finally `i` is assigned the value 12 and the for loop updates `sum` for the last time.
`r unhide()`

:::

::: {.webex-box .webex-check}

**Using the above code, what is the value of $1+2+3+...+12$?**
`r fitb(c("78", "78.0", "78.00", "78.000", "78.0000"))`

:::

You might not always know how many times you want to repeat certain lines of code. When this is the case the `while()` function can be used to repeat code while a given condition is satisfied. Once this condition is no longer `TRUE`, the loop will stop. The `while()` function has a similar setup to for loops.

```{r func4, eval=FALSE}
while(condition){
  statements
}
```

* `condition`: this is logical statement that can take the values `TRUE` or `FALSE`. When it is true, the statements will be repeated. When it is `FALSE`, the statements will not be evaluated.
* `statements`: these are the lines of code you want to repeat a number of times.

We can use the `while()` function to approximate the square root of a positive number, $x$. The sequence

$$
x_n = \frac{x_{n-1}}{2}+\frac{x}{2x_{n-1}}
$$

tends to $\sqrt{x}$ as $n\rightarrow\infty$.

Suppose we wanted to approximate $\sqrt{10,034}$ to within 0.0001. We can start with an initial guess of 10 and use the `while()` function to update this guess.

```{r func5}
x <- 10034
old_guess <- 10
new_guess <- (old_guess/2) + (x/(2*old_guess))
i <- 0
```

```{r func6}
while(abs(new_guess - old_guess) > 0.0001){
  old_guess <- new_guess
  new_guess <- (old_guess/2) + (x/(2*old_guess))
  i <- i+1
}
```

```{r func7}
c(new_guess, i)
```

We have used the value `i` here to track how many iterations the `while()` function runs for. We have also had to manually create the first `new_guess`, that is $x_1$. Note that the function `abs()` that is used within the `condition` returns the absolute value of the argument supplied. Look through this code carefully and make sure you understand what each part is doing.

'If statements' are another useful form of flow control. They have a very similar set up to the `for()` and `while()` functions.

```{r func8, eval=FALSE}
if(condition){
  statements
} else {
  statements
}
```

* `condition`: this is a logical statement that can take the values `TRUE` or `FALSE`. If the condition is `TRUE` then the  statements in the `{ }` immediately after will be evaluated.
* `else`: this is an optional part of the if statement. If there is code you wish to run when the condition is `FALSE`, it is placed in the `{ }` after this statement.

For example, we can write code to tell us whether a random value generated from the standard normal distribution lies in the region [-1, 1] using the following code.

```{r func9}
x <- rnorm(n = 1, mean = 0, sd = 1)

if(x >= -1 & x <= 1){
  print("x is in the region [-1, 1]")
} else {
  print("x is not in the region [-1, 1]")
}
```

::: {.webex-box .webex-check}

**What would the value of `y` be after running the following if statement? Try to answer without running the code yourself.**

```{r func-ans3}
x <- -4

if(x > 0){
  y <- x^2
} else {
  y <- -(x^2)
}
```

`r mcq(c("16", answer="-16", "-8", "8", "NA"))`

:::

<!-- ifelse() -->

---

You can read more about for loops and if statements in [Section 1.16 Flow Control](https://read.kortext.com/reader/pdf/92741/69) of *Probability and Statistics with R*.

## Creating functions

So far we have used a number of in built functions in R. It is also possible to create your own functions, specifying the required arguments and the what it does.

The setup for creating your own function using `function()` is as follows.

```{r func10, eval=FALSE}
fname <- function(argument1, argument2, ...){
  expression
}
```

* `fname`: this is the name you want to give to your function. It can be anything you choose but try to steer clear of using names of functions that already exist.
* `argument1`: this is the name of the first argument to be given when you use your function. You can add as many arguments as is necessary for your function to work.
* `expression`: this is what you want your function to evaluate. You write this out using the argument names you have specified.

For example, we can write a function, `normal()`, that:

* draws `n` random values from a normal distribution with mean `m` and standard deviation `s`
* orders them from smallest to largest
* for each random value $x$, returns the probability $\mathbb{P}(X\leq x)$ and the value of the probability density function at this value, $f(x)$

```{r func11}
normal <- function(n, m = 0, s = 1){
  x <- rnorm(n = n, mean = m, sd = s)
  x <- sort(x)
  prob <- pnorm(q = x, mean = m, sd = s)
  pdf <- dnorm(x = x, mean = m, sd = s)
  cbind(x, prob, pdf)
}
```

Here we have specified default values of `m = 0` and `s = 1` in our function. This means that if someone uses the function `normal` without specifying values for these arguments, then it will automatically use `m = 0` and `s = 1`.

Note that any vectors we create within the expression for the function are not saved to your **Environment** tab.

We can then use this new function `normal()` to see this information for 5 random values from the $N(200, 40)$ distribution.

```{r func12, echo=c(2)}
set.seed(1406)
normal(n = 5, m = 200, s = 40)
```

The arguments you use within a function can be either 'named' or 'positional'.

Named arguments are ones where you use the name given in the setup of the function. For example, in our function `normal()`, the argument for the mean is called `m`. You can change the value of the mean using `m = 4` within the `normal()` function for example.

Positional arguments are ones where you don't use the name of the argument given in the setup of the function. For example, we could replicate the above output using the following code where the argument names have not been used.

```{r func13, echo=c(2)}
set.seed(1406)
normal(5, 200, 40)
```

The important thing to remember when using positional arguments is that they are in the same order that their names are specified in the setup of the function so R knows how to match up the values correctly. For example, in the `normal()` function, the mean needs to be the second argument specified.

---

For more information on writing your own functions in R, see [Section 1.17 Creating Functions](https://read.kortext.com/reader/pdf/92741/74) in *Probability and Statistics with R*.











