---
title: "Welcome!"
number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, comment=NA)
library(webexercises)
library(PASWR2)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyverse)
```

# Introduction {#sec-introduction}

Welcome to the ninth S2S lab. Today we will complete the statistical report that we started in [Lab 8](https://moodle.gla.ac.uk/pluginfile.php/9502911/mod_resource/content/2/){target="_blank"} by completing the **Statistical Analysis** and **Conclusion** sections. If you have completed Lab 8, you should have the beginnings of a statistical analysis report saved.

We will take a look at different ways of calculating **confidence intervals** in different scenarios in R, which will allow us to complete the statistical report.

<!-- question callout example
:::{.question}
::::{.question-header}
Question
::::
::::{.question-container}
This is where the question text would go.
::::
:::
-->

<!--
::: panel-tabset
## R Script
```{r, eval = FALSE}

```

## Console
```{r, echo = FALSE}

```
:::
-->










