# Summary of Functions

## [Barplots](01-qualitative.qmd#sec-bars) {.unnumbered}

```{r, eval = FALSE}
barplot(height = ,
        legend.text = ,
        beside = FALSE,
        col = ,
        main = ,
        xlab = ,
        ylab = ,
        ylim = )
```

## [Dot Charts](01-qualitative.qmd#sec-dots) {.unnumbered}

```{r, eval = FALSE}
dotchart(x = ,
         bg = ,
         main = ,
         xlab = ,
         ylab = )
```

## [Pie Charts](01-qualitative.qmd#sec-pies) {.unnumbered}

```{r, eval = FALSE}
pie(x = ,
    radius = ,
    col = ,
    main = )
```

## [Stem and Leaf Plots](02-quantitative.qmd#sec-stem) {.unnumbered}

```{r, eval = FALSE}
stem(x = , scale = 1)
```

## [Histograms](02-quantitative.qmd#sec-hist) {.unnumbered}

```{r, eval = FALSE}
hist(x = ,
     breaks = ,
     freq = TRUE,
     right = TRUE,
     include.lowest = TRUE,
     col = ,
     main = ,
     xlab = ,
     ylab = ,
     ylim = )
```

## [Kernel Density Estimators](02-quantitative.qmd#sec-kernel) {.unnumbered}

```{r, eval = FALSE}
density(x)
```

```{r, eval = FALSE}
lines(col = , lwd = )
```

```{r, eval = FALSE}
plot(col = ,
     lwd = ,
     main = ,
     xlab = ,
     ylab = ,)
```

## [QQ Plots](03-compare.qmd#sec-qq) {.unnumbered}

```{r, eval = FALSE}
qqnorm(y = , main = )
```

```{r, eval = FALSE}
qqline(y = )
```





