# Summary of Functions

## [Packages](01-packages.qmd#sec-packages) {.unnumbered}

```{r, eval=FALSE}
install.packages("")
```

```{r, eval=FALSE}
library()
```

## [Arrays](02-data.qmd#sec-arrays) {.unnumbered}

```{r, eval=FALSE}
array(data = , dim = )
```

## [Creating matrices](02-data.qmd#sec-creating-mat) {.unnumbered}

```{r, eval=FALSE}
matrix(data = , nrow = , ncol = , byrow = FALSE)
```

<!--
```{r, eval=FALSE}
dim()
```
-->

## [Naming rows and columns](02-data.qmd#sec-naming-mat) {.unnumbered}

```{r, eval=FALSE}
dimnames()
```

## [Calculating statistics](02-data.qmd#sec-calculate-mat) {.unnumbered}

```{r, eval=FALSE}
apply(X = , MARGIN = , FUN = )
```

<!--
## [Vector/Matrix multiplication](02-data.qmd#sec-mat-mult) {.unnumbered}

```{r, eval=FALSE}
t()
```

```{r, eval=FALSE}
solve()
```
-->

## [Factors](02-data.qmd#sec-fact) {.unnumbered}

```{r, eval=FALSE}
factor(x = , levels = , labels = )
```

## [Data frames](02-data.qmd#sec-df) {.unnumbered}

```{r, eval=FALSE}
data.frame(stringsAsFactors = , row.names = )
```

```{r, eval=FALSE}
str()
```

```{r, eval=FALSE}
attach()
```

```{r, eval=FALSE}
detach()
```

```{r, eval=FALSE}
data()
```

```{r, eval=FALSE}
head()
```
